Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals

### 2015-10-13 manuscript/supplemental_material.Rnw
add supplemental material

\begin{document}
\author[1,2]{Michael R. McGowen}
\author[1]{Don Armstrong}
\author[1]{Amy Weckle}
\author[1]{Jason Caravas}
\author[3]{Dalen Agnew}
\author[4]{Kurt Benirschke}
\author[5]{Sue Savage-Rumbaugh}
\author[6,7]{Chong Jai Kim}
\author[8]{Günter P. Wagner}
\author[6]{Roberto Romero}
\author[1,6,9,*]{Derek E. Wildman}
\title{The evolution of gene expression in the term placenta of viviparous mammals}
% These are all wrong, but we'll change them later.
\affil[1]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
\affil[2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
\affil[3]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
\affil[4]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
\affil[5]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
\affil[6]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
\affil[7]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
\affil[8]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
\affil[9]{Department of Obstetrics and Gynecology, Wayne State University School of Medicine, Detroit, MI, USA}
\affil[*]{To whom correspondence should be addressed}
\date{\today}

% Acronyms
\input{acronyms}
% \twocolumn
\maketitle
\onehalfspacing

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -2, digits = 2)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ape")
library("ggtree")

@@

<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset")))>>=
for (a in paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset"))) {
    load(a)
}
setkey(genes.to.tree,"gene_id")
star.logs[,value:=gsub("%$","",value)]
star.logs.wide <- 
    data.table(dcast(star.logs,species+file~field))
setnames(star.logs.wide,
         "Average input read length",
         "input.read.length")
setnames(star.logs.wide,
         "Number of input reads",
         "input.reads")
setnames(star.logs.wide,
         "Uniquely mapped reads number",
         "unique.map.reads")
@

\section{Abstract}

The placenta is arguably the most anatomically variable organ in
mammals. Understanding gene expression patterns among placentas from
multiple species can illuminate the molecular basis of placenta
function as well as potentially enrich our knowledge of the biology of
obstetrical syndromes. We profiled the term placenta by using Illumina
RNA-Seq technology to sequence the transcriptomes of
\Sexpr{star.logs[,length(unique(species))]} species of therian (i.e.
marsupials and “placental”) mammals. We identified significant shifts
in gene expression along specific evolutionary lineages using a model
of character evolution. Genes that showed significant elevation in
expression along the human lineage ($\mathrm{FDR}\le 0.01$) were
enriched in GO categories such as angiogenesis, tube morphogenesis,
cell migration, and regulation of inflammatory response. Three of
these genes (KISS1, ADAM12, GPC3) showed exceptionally high expression
in humans; these genes are implicated in preeclampsia and may indicate
that the risk of preeclampsia recently emerged on the human lineage.

\textbf{Keywords:} mammal, gene expression, evolution, RNA-seq, placenta

\section{Introduction}

The main role of the placenta in viviparous mammals is the transfer of
nutrients, oxygen, and waste between mother and offspring. Although
these functions are conserved among mammals, the placenta is perhaps
the most variable organ in terms of structure and form
\cite{Mossman.1937:comparativemorphogenesisofthe}. Species differ in
the shape of attachment (i.e., discoid, diffuse, cotyledonary,
zonary), degree of invasiveness of fetal tissues into maternal tissues
(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
interdigitation of maternal-fetal tissue (i.e., villous, folded,
trabecular, lamellar, labrynthine)
\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
mammals differ drastically in the number and size of their offspring,
as well as the length of gestation, traits often associated with
specific aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}.

The placenta arose at some point before the origin of therian mammals,
a group that includes both marsupials and eutherian or “placental”
mammals. The term “placental mammals” is misleading, as marsupials
also possess a placenta, although its direct apposition to maternal
tissues lasts for a proportionally shorter time during gestation
\cite{Renfree2010:ReviewMarsupialsplacentalmammals}. Marsupials also differ from eutherian mammals in
that most possess only a choriovitelline placenta, produced from the
fusion of the chorion and yolk sac, during most of gestation. Although
some eutherian mammals may possess a choriovitelline placenta at some
point in development, the major site of nutrient exchange in
placentals occurs via the chorioallantoic placenta, produced from the
fusion of the chorion and allantois \cite{Mossman.1987:vertebratefetalmembranes}. It is unclear
whether the first placenta was more like that of marsupials or
eutherians; however, using phylogenetic methods, the ancestral
eutherian was reconstructed as possessing a relatively invasive
hemochorial placenta, in which the maternal and fetal blood supply is
separated by one tissue layer, as in humans and mouse
\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.

Although the morphology of the mammalian placenta has been well
characterized in a number of species \cite{Mossman.1987:vertebratefetalmembranes}, the diversity of
the molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}. Studies of certain model species
(i.e., humans, mice, cattle) have revealed a stunning divergence in
the genetics of the placenta, especially in later stages of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive
placenta-specific gene families with expression patterns that can vary
in space and time \cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}. These
include independent expansions of the prolactins in mouse and cow,
pregnancy-specific glycoproteins (PSGs) in humans and mouse,
pregnancy-associated glycoproteins (PAGs) in cow, and the expansion of
the growth hormone and chorionic gonadotropin beta genes in hominoid
primates \cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}. Expression patterns of
specific genes also can vary widely between species. For example, the
expression of orthologous genes in the placenta have been found to
vary substantially between humans and mice \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of SIGLEC6 to placental expression in humans differs from
both chimpanzees and mice \cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

One explanation for the divergent paths of the placental interface is
the distinct evolutionary interplay between maternal and fetal
strategies during pregnancy \cite{Haig1993:Geneticconflictsinhuman}. The maternal strategy is to
requisition an adequate degree of nutrients to the fetus without
risking the health of the mother, while the fetal strategy via
paternal genes is to extract as many resources from the mother as
possible. Therefore, an interplay between these extremes can set up an
intricate “arms race”, involving the imprinting of parental alleles,
production of new genes, and enlistment of existing genes to
expression at the maternal-fetal interface \cite{Haig1993:Geneticconflictsinhuman}.

One symptom
of this maternal-fetal conflict is its potential affect on human
health \cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. 
In addition to having an invasive
hemochorial placenta, humans possess a deeply invasive trophoblast
layer that can penetrate the myometrium of the uterus 
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
 Defects in placental growth are known to be
associated with the devastating condition of preeclampsia, which can
cause hypertension, proteinuria, and has the potential to lead to
death of the mother and/or infant \DLA{(citation)}. Although cases have been
documented in chimpanzees and gorillas \DLA{(citation)}, preeclampsia has a
far higher incidence in human pregnancies with as much as 4\% of
pregnancies being affected in the United States \DLA{(citation)}.
Chimpanzees and gorillas also have deeply invasive trophoblasts,
although the progression of invasion in great apes may be slower than
humans \cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however, large-scale expression
studies of great ape placentas have not been conducted. The increasing
affordability of next-generation sequencing technology has enabled the
large-scale study of specific tissues from a wide array of species to
address broader phylogenetic questions in the evolution of gene
expression \cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use Illumina RNA-Seq
technology to sequence and quantify the mRNA environment of six
mammalian placentas at term, including that of the human, bonobo (the
sister species to the chimpanzee), dog, cow, African elephant, and
opossum, representing marsupials as well as three major branches of
eutherian mammals (Euarchontoglires, Laurasiatheria, Afrotheria).
These placentas also represent a wide array of placental morphologies
as well as traits such as number and size of offspring and length of
gestation \DLA{(Table 1)}. We use this data, along with existing RNA-Seq
data from mouse \cite{Wang.ea2011:surveyfornovelimprinted}, to address the major differences
between expression in marsupial and eutherian mammals at term,
identify the evolution of expression in important placental genes such
as hormones and galectins, identify novel lineage-specific expression
patterns, and quantify human specific differences in gene expression
that may be related to obstetrical disorders such as preeclampsia.

\section{Methods}

\subsection{Collection of Placental Tissue and RNA Extraction}

Fetal tissue was collected from the placenta of six mammalian species
(\textit{Homo sapiens} [human], \textit{Pan paniscus} [bonobo], \textit{Canis familiaris}
[domestic dog], \textit{Bos taurus} [domestic cow], \textit{Loxodonta africana} [African
savanna elephant], \textit{Monodelphis domestica} [gray short-tailed
opossum], \textit{Nannospalax gallili} [?], \textit{Spalax carmelli} [?]).
Information on specimen ID, place and time of collection,
and collector is shown for each sample in \DLA{Supplemental Table 1}. In the
case of \textit{Pan}, \textit{Bos}, and \textit{Loxodonta}, the placenta was collected at term
upon expulsion from the mother following delivery. Placental tissue
from \textit{Homo sapiens} was collected upon delivery via a caesarian section.
Near-term placental tissue from \textit{Canis familiaris} was collected upon
the death of the mother from a car accident. Placental tissue from
\textit{Monodelphis domestica} was sampled at fetal stage 32, roughly one day
prior to birth. Only the fetal portion of the placenta was sampled at
the maternal-fetal interface (i.e. villous portion for \textit{Homo}, \textit{Pan},
\textit{Canis}, and \textit{Loxodonta}; cotyledon for \textit{Bos}; diffuse placenta for
\textit{Monodelphis}). All tissues were collected and stored in RNAlater (Life
Technologies, Carlsbad, CA) until RNA isolation could be performed.
Tissues were homogenized in Trizol, and total RNA was isolated
following the manufacturer’s instructions. RNA was further purified
using RNeasy Midi Kit (Qiagen, Valencia, CA), in conjunction with the
RNase-Free DNase Set (Qiagen, Valencia, CA), following the
manufacturer’s protocol to remove any contaminating DNA from the
sample. Total RNA was then evaluated on a Nanodrop 1000 (Thermo
Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100 (Santa
Clara, CA) to confirm concentration and quality of the isolation. RNA
quality was assessed using A260/A280 and A260/A230 absorbance values.

\DLA{Add details for new sequences and fix numbers}

\subsection{Sequencing}

\DLA{Fix how the sequencing was done}

All RNA-Seq mRNA isolation, library construction, and Illumina
sequencing was performed by the Applied Genomics Technology Center
(AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
total RNA and standard Illumina protocol. Paired-end sequencing was
performed using the Illumina Genome Analyzer II with an insert size of
230 base pairs (bp). \textit{Bos}, \textit{Loxodonta}, and
\textit{Monodelphis} were sequenced with read lengths of 75 bp;
\textit{Canis}, \textit{Homo}, and \textit{Pan} were sequenced with
read lengths of 150 bp.

For comparison, additional sequences of placenta in \textit{Mus
  musculus} \cite{Wang.ea2011:surveyfornovelimprinted} and
\textit{Homo sapiens}
\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} were obtained
from the \ac{SRA}.

\subsection{Transcriptome Assembly, Alignment, and Quantification}

The quality of reads was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (Reference of ---
in \fref{tab:ref_genome_aligned_species}) were assembled into a
reference using Trinity v2014-07-17
\cite{Haas.ea2013:Denovotranscriptsequence} and annotated using
Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against either
the \textit{Homo sapiens} or \textit{Mus musculus} transcriptome
(\fref{tab:ref_genome_aligned_species}).
% 
All transcriptomes were then aligned to the appropriate reference
genome using STAR v2.4.2a
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
values in \ac{FPKM}
\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
%
The genome of the closely-related \textit{Pan troglodytes} was used as
a reference for \textit{Pan paniscus}. The separation between the two
recognized Pan species is generally perceived to be $\approx 1$ Mya
\cite{Prüfer.ea2012:bonobogenomecomparedwith}.

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Common Name        & Reference & Read Length  \\
    \midrule
    \textit{Ateles fusciceps}     & Spider monkey      & \textit{Homo sapiens} & 100 \\
    \textit{Bos taurus}           & Cow                & UMD3.1     & 76 \\
    \textit{Canis familiaris}     & Dog                & CanFam3.1  & 151 \\
    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & 101 \\
    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & 76 \\
    \textit{Homo sapiens}         & Human              & GRCh38     & 76, 151 \\
    \textit{Monodelphis domestica} & Opossum           & BROAD05    & 76 \\
    \textit{Mus musculus}         & Mouse              & GRCm38     & 36,44,151 \\
    \textit{Nannospalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 151 \\
    \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \bottomrule
  \end{tabular}
  \caption{Reference genomes for aligned species}
  \label{tab:ref_genome_aligned_species}
\end{table}

\subsection{Exploration of Individual Transcriptome Data}

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using DAVID
\cite{Huang.ea2009:Systematicandintegrativeanalysis,Sherman.ea2007:DAVIDKnowledgebasegene-centereddatabase}.
DAVID ranks functional annotation clusters according to enrichment
score (ES); functional annotation clusters are considered
significantly enriched if ES>1.3, as this corresponds to p<0.05
\cite{Huang.ea2009:Systematicandintegrativeanalysis}. We executed
analyses with two gene lists per transcriptome: 1) annotated genes
with expression levels of the top 5\% of the frequency distribution of
\ac{FPKM} and 2) the top 100 genes with the highest \ac{FPKM}. Due to the
paucity of annotated gene ontology terms for some species, gene lists
were compared using gene ontology terms derived from the human genome.

\subsection{Comparative Analysis of Gene Expression}

All 1:1 orthologues present among human, bonobo, mouse, dog, cow,
elephant, and opossum were determined using Ensembl v.64 gene trees
via the Biomart online data management
tool\cite{Kasprzyk2011:BioMartdrivingparadigmchange}. A total of 6882
1:1 orthologous genes were expressed in at least one species. For
further analysis, we eliminated all genes that did not have at least
one species with an $\mathrm{FPKM}\ge10$. A list of these 4259 1:1 orthologous
genes was retained for future analyses of gene expression evolution
described below.

We used the software OUCH (Ornstein–Uhlenbeck for Comparative
Hypotheses) in the R statistical environment to investigate
significant shifts in gene expression levels over
time\cite{Butler.2004:phlogeneticcomparativeanalysis}. OUCH institutes
an Ornstein-Uhlenbeck (OU) model of evolution, which allows for
different optima for quantitative trait values along user-specified
branches of a phylogenetic tree. The OU model explicitly accounts for
change in quantitative trait values by introducing two parameters into
the model of stochastic Brownian motion of change in trait value along
a phylogenetic tree through time. These parameters include, $\theta$,
or an optimum trait value, and $\alpha$, or the strength of selection
(interpreted as “pull” toward the optimum). In addition, the OU model
can also integrate evolutionary change by assuming that each
user-specified set of braches has its own separate OU parameters
\cite{Butler.2004:phlogeneticcomparativeanalysis}.

Various OU branch models were tested against a model in which no
variation in optima among branches was allowed. The following branches
were specified in separate models to have a different optimum than all
other branches: human, bonobo, mouse, dog, cow, elephant, opossum,
human+bonobo, Euarchontoglires, and Boreoeutheria. Likelihood values
were computed for each branch model and compared to an OU model with
one optimum using likelihood ratio tests with df=1. We filtered out
genes with universally low expression levels, and included only genes
in which expression in at least one species had an
$\mathrm{FPKM}\ge10$ as described above. We corrected for multiple
testing using the false discovery rate (FDR) and calculated adjusted
$p$ values (q-values) using the Benjamini-Hochberg
correction\cite{Benjamini1995:bh_procedure}. We used the
time-calibrated phylogenetic tree of
\cite{Meredith.ea2011:ImpactsofCretaceousTerrestrial} with all species
not used in this analyses pruned.

To determine enrichment of functional categories in lists of genes
that display a significant shift in expression on a specific branch,
we performed functional annotation cluster analyses on each branch
using DAVID
\cite{Huang.ea2009:Systematicandintegrativeanalysis,Sherman.ea2007:DAVIDKnowledgebasegene-centereddatabase}.
DAVID ranks functional annotation clusters according to enrichment
score (ES); functional annotation clusters are considered
significantly enriched if $\mathrm{ES}\ge1.3$, as this corresponds to
$p\le0.05$ \cite{Huang.ea2009:Systematicandintegrativeanalysis}. As
above, due to the paucity of annotated gene ontology terms for some
species, gene lists were compared using gene ontology terms derived
from the human genome.

\subsection{Correlation of Tissue Specific Index with Lineage Specific Genes}

To evaluate the extent of tissue specificity of each gene, we
calculated the tissue specificity index based on human microarray data
from (Su et al., 2004) available via BioGPS. The tissue specificity
index (TSI) measures the extent to which a specific gene has
predominant expression in only one tissue (Yanai et al., 2005). A TSI
of 1.0 would signify that a gene was solely expressed in only one
tissue, while a TSI of 0.0 would signify that a gene had equal
expression in all tissues evaluated. We used the formula available in
Yanai et al. (2005) to calculate the TSI, and identified the top four
tissues which expressed each probe set from (Su et al., 2004).

\subsection{Validation of Primate Expression using Spider Monkey Placenta Transcriptome}

To verify expression levels in human and bonobo, we also generated
RNA-Seq data for the placenta transcriptome of the black-headed spider
monkey (\textit{Ateles fusciceps}) using similar protocols discussed above.
Ateles is outside the clade, which includes hominids and can be used
to polarize expression values of highly expressed genes in Homo and
Pan. We sampled placental tissue from an Ateles placenta expelled from
the mother shortly after birth (see Supplemental Table X). We used
similar procedures discussed above to extract and sequence this
sample. [DEREK—I’m unsure how to proceed on describing this so as to
not step on Natalie’s toes]

\DLA{Complete discussion of the ateles placenta}

\section{Results}

\subsection{Transcriptome Statistics}

@@

For transcriptomes sequenced with 75 bp, %
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with at least 96\% of reads \DLA{This is wrong}
mapping to the respective reference genome (\fref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with at least 99\% of reads mapping to the
respective reference genome \DLA{This is also wrong}.
%
%
For transcriptomes sequenced with 35 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with at least 99\% of reads mapping to the
respective reference genome \DLA{This is also wrong}.
%
All reads were submitted to the NCBI Sequence Read Archive (SRA)
(Accession Number XXX-YYY).

\DLA{Need to do an embargoed submission to SRA}

The percentage of bases that mapped to Ensembl annotated mRNA ranged
from ~35\% in elephant to ~85\% in human, with percentages of bases
mapping to annotated protein-coding segments of genes ranging from
~34\% (elephant) to ~54\% (human), untranslated regions (UTRs) ranging
from ~0.6\% (elephant) to ~31\% (human), and known introns ranging
from ~5.5\% (dog) to 11\% (bonobo). Median percent genomic coverage
ranged from ~43\% in elephant to 80\% in bonobo. The number of
annotated transcripts to which reads mapped in each respective genomes
ranged from 4119 (bonobo) to 6506 (human) for transcripts with \ac{FPKM}>10
and 11689 (dog) to 30733 (human) for transcripts with \ac{FPKM}>1. The
greater degree of mapping to annotated protein-coding sequences in
humans corresponds to the greater number of Ensembl annotated
sequences in the human genome compared to the other five reference
genomes (Flicek, 2012).

\DLA{Summarize all of this in a table}

The top 50 expressed genes in each genome analyzed are shown in
\DLA{Tables 2-7}, and all genes for each transcriptome are listed in
\DLA{Supplementary Tables X-XX} [These are large Excel files to be
included with the Supplemental Material]. In addition, specific genes
including placental hormones and hormone-related genes, galectins, and
insulin-like growth factor related genes are shown with \ac{FPKM} values
side to side for all transcriptomes including the reanalysis of mouse
from \cite{Wang.ea2011:surveyfornovelimprinted} in \DLA{Table 8}.
Below we describe each transcriptome separately, specifically
concentrating on highly-expressed genes.

<<top.genes.function,echo=FALSE,results="hide">>=
top.genes.table <- function(species.name,n=50) {
    temp <- 
        combined.fpkm[species==species.name &
                          !(human_name %in% housekeeping.genes.superset[,gene_short_name]) &
                              !grepl("(RNA\\d|MT)-",human_name)
                      ][order(-mean_fpkm)][!duplicated(gene_short_name)][1:n,list(gene_short_name,
                                                                                  human_name,mean_fpkm)]
    setnames(temp,c("Gene name","Human Name","FPKM"))
    if (species.name=="homo sapiens") {
        temp <- temp[,-2,with=FALSE]
    }
    return(temp)
}
gene.expression <- function(species.name,gene) {
    combined.fpkm[species==species.name &
                      gene_short_name==gene,mean_fpkm]
}

\subsection{Human Placenta Transcriptome}

\begin{wraptable}{R}{0.3\textwidth}
  \tiny
<<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
top.human.genes <- top.genes.table(species.name="homo sapiens")
print(xtable(top.human.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.human.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Homo sapiens}}
\label{tab:top50_human_placenta}
\end{wraptable}

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \textit{Homo sapiens} are shown in
\fref{tab:top50_human_placenta} with a list of all genes in
Supplemental Table X. The gene with by far the highest \ac{FPKM} value
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
in humans is \ac{CSH1}, followed by \ac{CSH2}, and \ac{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within hominoids. Another member of the growth
hormone gene family, \ac{GH2} (\ac{FPKM}=%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the human placenta
(\fref{tab:top50_human_placenta}).
% 
Ten pregnancy-specific beta-1-glycoprotein genes are highly-expressed
(\ac{FPKM}>100) in the term placenta of humans, with \ac{PSG4} showing the
greatest degree of expression %
(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG4")}). % 
These
genes are members of a hominoid-specific expansion of the
carcinoembryonic antigen (CEA) family and known to be expressed by the
syncytiotrophoblast (Lei et al., 1992; Rand and Cross, 2008). Genes
involved in the production of estrogen (CYP19A1) and progesterone
(HSD3B1) show exceptionally high expression in the human placenta
(\ac{FPKM} = 4817.15 and 829.846 respectively), as well as CGA
(\ac{FPKM}= 4793.08), which generates the alpha subnit of multiple
human chorionic gonadotropin hormones (Table 2). The genes producing
the beta subunits of the chorionic gonadotropins (CGB, CGB1, CGB2,
CGB5, CGB7, CGB8) are either not expressed (CGB8) or expressed
modestly (FPKM≤20.826; Table 8). Other highly expressed genes in the
top 50 include members of the hemoglobin family (e.g. HBA1, HBB, HBG2,
HBA2), the maternally imprinted long noncoding RNA gene H19, genes
coding for proteins required for growth and/or cell proliferation
(ADAM12, KISS1, IGF2, GPC3, CSF3R), anticoagulation (TFPI2), and
glucose transport (SLC2A1) among others (Table 2, Supplemental Table
X)

\subsection{Bonobo Placenta Transcriptome}

\begin{wraptable}{L}{0.3\textwidth}
  \tiny
<<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
print(xtable(top.bonobo.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.bonobo.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Pan paniscus}}
\label{tab:top50_bonobo_placenta}
\end{wraptable}

The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
the bonobo are shown in \fref{tab:top50_bonobo_placenta} with a list of all genes in
Supplemental Table X. Expression in the bonobo placenta differs
somewhat from that of the human, as evidenced by the makeup of the top
50 genes (Table 3). The highest expression (\ac{FPKM}=%
\Sexpr{gene.expression("pan paniscus","PRG2")}%
%
) in bonobo is \ac{PRG2}, which is involved in insulin
growth factor signaling and angiogenesis during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}. \ac{PRG2} is also expressed at relatively high levels in the
human placenta (\ac{FPKM}=\Sexpr{gene.expression("homo sapiens","PRG2")}; Supplemental Table X), although at
comparatively much lower levels than the bonobo. A bonobo gene closely
related to the chorionic somatomammotropins (Q95MK7\_PANTR) is also
expressed at high levels (\ac{FPKM}=1232.56; \fref{tab:top50_bonobo_placenta}). Two other Pan genes
included within this family (ENSPTRG00000030805 which shows similarity
to GH2 and ENSPTRG00000034560) show expression levels of \ac{FPKM}>89. Six
pregnancy-specific beta-1-glycoprotein genes were expressed in the
bonobo placenta with \ac{FPKM} ranging from 20.885-167.961 (Supplemental
Table X). Placental expression of the genes involved in estrogen and
progesterone production (CYP19A1, HSD3B1 respectively) show lower
expression in the bonobo as compared to humans (CYP19A1 \ac{FPKM}= 48.6877;
HSD3B1 \ac{FPKM}= 54.5301; Table 8). In addition, CGA is expressed in the
bonobo placenta (\ac{FPKM}=202.629), but at much lower levels than humans
as well (Table 8). Two Ensembl-annotated beta subunits of the
chorionic gonadotropin genes are expressed in bonobo, one related to
CGB and another related to CGB7; both are not highly expressed (Table
8). Among the other top 50 expressed genes, there is high expression
of ribosomal proteins (e.g., RPL18, RPL23), keratins (e.g., KRT7,
KRT19), heat-shock proteins (e.g., HSPA5, HSPA1A), serpins (SERPINE1,
SERPINE2), insulin growth factor 2 (IGF2) and the hormone
adrenomedullin (ADM).

\subsection{Spider Monkey Transcriptome}

\begin{wraptable}{R}{0.3\textwidth}
  \tiny
<<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
print(xtable(top.spider.monkey.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.spider.monkey.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Ateles fusciceps}}
\label{tab:top50_spider_monkey_placenta}
\end{wraptable}

\subsection{Dog Placenta Transcriptome}

\begin{wraptable}{R}{0.3\textwidth}
  \tiny
<<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
top.dog.genes <- top.genes.table(species.name="canis familiaris")
print(xtable(top.dog.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.dog.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Canis familiaris}}
\label{tab:top50_dog_placenta}
\end{wraptable}

>>>>

\begin{document}

\makeatletter 
\renewcommand{\thefigure}{S\@arabic\c@figure} 
\makeatother

\makeatletter 
\renewcommand{\thetable}{S\@arabic\c@table} 
\makeatother

\title{Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals}

% \maketitle

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -2, digits = 2)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ape")
library("ggtree")
library("topGO")
library("org.Hs.eg.db")

@@

<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","palcenta_go_results_mf")))>>=
for (a in paste0("../data/",c("combined_fpkm","gene_trees",
                              "housekeeping_genes_superset",
                              "placenta_core_transcriptome",
                              "placenta_go_results_bp",
                              "placenta_go_results_mf"
                              ))) {
    load(a)
}
setkey(genes.to.tree,"gene_id")
star.logs[,value:=gsub("%$","",value)]
star.logs.wide <- 
    data.table(dcast(star.logs,species+file~field))
setnames(star.logs.wide,
         "Average input read length",
         "input.read.length")
setnames(star.logs.wide,
         "Number of input reads",
         "input.reads")
setnames(star.logs.wide,
         "Uniquely mapped reads number",
         "unique.map.reads")
@

@@

<<genes.expression,echo=FALSE,results="hide">>=
genes.expression.table <- function(species.name,min.fpkm=50) {
    temp <- 
        combined.fpkm[species==species.name][order(-mean_fpkm)][,list(name_or_id,
                          human_name,mean_fpkm)]
    temp <- temp[!duplicated(name_or_id)][mean_fpkm>=min.fpkm]
    setnames(temp,c("Gene name","Human Name","FPKM"))
    if (species.name=="homo sapiens") {
        temp <- temp[,-2,with=FALSE]
    }
    return(temp)
}
@

<<human.expressed.genes,echo=FALSE,results="asis">>=
print(xtable(genes.expression.table(species.name="homo sapiens"),
             caption="Expression of all genes with FPKM $\\ge 50$ in \\textit{Homo sapiens}",
             label="tab:all_human_placenta"
             ),
      booktabs=TRUE,
      tabular.environment="longtable",
      floating=FALSE,
      caption.placement="top",
      add.to.row = list(pos = list(0),command = "\\midrule \\endhead "),
      size="\\tiny"
      )
@
